Comparison of Fast Imaging Architectures for Large, Multi-scale, Hierarchical Aperture Arrays

### 2024-06-06 ms.tex
Update on Overleaf.

% \author{S. Author}
% \affiliation{Second Division, Organization, City, Pincode, State, Country}
% % \alsoaffiliation{Joint first authors}

% \author{T. Author}
% \affiliation{Second Division, Organization, City, Pincode, State, Country}

% \author{F.T. Author}
% \affiliation{Fourth Division, Organization, City, Pincode, State, Country}

@@

\cite{TMS2017}
\cite{Otobe+1994,Tegmark+2009,Tegmark+2010,Zheng+2014}
\cite{Morales2011,Thyagarajan+2017}
\cite{Cornwell+2008}
\cite{Beardsley+2017}

@@

@@

\section{Multi-scale, Hierarchical Aperture Arrays} \label{sec:multi-scale-arrays}

The specific aperture array examples we consider are the \texttt{SKA-low}, core of the SKA-low (\texttt{SKA-low-core}), low frequency Australian Very Long Baseline Array (\texttt{Aus-VLBA-low}), and low frequency inter-continental VLBA (\texttt{IC-VLBA-low}). The last two are hypothetical interferometers. It is assumed that $N_{\textrm{a/s}_m}=N_\textrm{a/s}=256$ (same for each station), $D_\textrm{a}/\lambda=1$, and $D_\textrm{s}/\lambda=17.5$ for each of these arrays, where, the nominal wavelength chosen here is $\lambda=2$~m. The corresponding station filling factor, $f_\textrm{s}=N_\textrm{a/s}\,(D_\textrm{a}/D_\textrm{s})^2$, is 0.836 for all the arrays. ~\texttt{SKA-low-core}, ~\texttt{SKA-low}, ~\texttt{Aus-VLBA-low}, and ~\texttt{IC-VLBA-low} are taken to consist of $N_\textrm{s}=256, 512, 512,$ and 512 stations in an array diameter of $D_\textrm{A}/\lambda=525, 3.25\times 10^4, 1.95\times 10^6$, and $3.9\times 10^6$, respectively. The corresponding array filling factors, defined as $f_\textrm{A}=N_\textrm{s}\,(D_\textrm{s}/D_\textrm{A})^2$, are 0.284, $1.48\times 10^{-4}$, $4.12\times 10^{-8}$, and $1.03\times 10^{-8}$, respectively. These parameters for the chosen examples are summarised in Table~\ref{tab:array_params}.

@@

Fast imaging involves processing data on multiple time scales. The channelisation of a voltage time stream of length $\delta t$ with a time resolution of $t_\textrm{s}$ will correspond to spectral channels of width $\delta\nu=1/\delta t$ and a bandwidth of $B=1/t_\textrm{s}$. The number of spectral channels is $N_\nu=B/\delta\nu=\delta t/t_\textrm{s}$. Images are expected to be produced at a cadence of $t_\textrm{acc}$, which is typically larger than $\delta t$ and smaller than the time scale on which calibration needs to be updated. In this paper, the nominal values chosen for $\delta t$ is 25~$\mu$s, and we explore the possibilities for $t_\textrm{acc}$ ranging from 0.1~ms to 10~s.

@@

@@

Figure~\ref{fig:1D-coherent-compcost} shows a comparison of the compute cost densities in the discovery phase space of different multi-scale imaging architectures combining four options -- VBF, EPIC, XBF, and XFFT -- on the inter-station data with two options -- VBF and EPIC -- on the intra-station data for various array parameters and imaging cadences.

\begin{figure*}
\centering
\subfloat[][\texttt{LAMBDA-I} \label{fig:1D-coherent-compcost-LAMBDA-I}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_LAMBDA-I.pdf}
} \\
\subfloat[][\texttt{SKA1-Low-core} \label{fig:1D-coherent-compcost-LAMBDA-I}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_SKA1-low-core.pdf}
} \\
\subfloat[][\texttt{SKA1-Low} \label{fig:1D-coherent-compcost-LAMBDA-I}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_SKA1-low.pdf}
} 
\caption{One-dimensional slices of the computational cost density for (a) \texttt{LAMBDA-I}, (b) \texttt{SKA-low-core}, and (c) \texttt{SKA-low} using  coherent imaging at intra- and inter-station levels via VBF (gray) and EPIC (black) for intra-station processing, and VBF (dotted), EPIC (solid), XBF (dashed), and XFFT (dot-dashed) for inter-station processing of data. \label{fig:1D-incoherent-compcost-a}}
\end{figure*}

\begin{figure*}\ContinuedFloat
\subfloat[][\texttt{CASPA} \label{fig:1D-coherent-compcost-CASPA}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_CASPA.pdf}
} \\
\subfloat[][\texttt{FarView} \label{fig:1D-coherent-compcost-FarView}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_FarView.pdf}
}
\caption{Contd. Same as above but for (d) \texttt{CASPA}, and (e) \texttt{FarView} \label{fig:1D-coherent-compcost}}
\end{figure*}

@@

The inter-station processing, however, varies significantly due to the diversity in the various inter-station array parameters and imaging cadence considered. For example, 
\begin{itemize}
    \item \texttt{LAMBDA-I} will benefit from deploying the correlator beamforming (XBF) architecture for inter-station processing. 
    \item \texttt{SKA1-low-core} will benefit from EPIC for the chosen parameters. However, if the number of stations, $N_\textrm{s}$, decreases,  the array diameter, $D_\textrm{A}$, increases or if the imaging cadence is slowed to $t_\textrm{acc} \gtrsim 5$~ms, the XFFT architecture starts to have an advantage over EPIC on the inter-station scale of data processing. 
    \item \texttt{SKA1-low} will benefit from the XFFT architecture for inter-station processing.
    \item \texttt{CASPA} will benefit from using the XBF architecture for inter-station processing due to significantly small number of stations.
    \item \texttt{FarView} appears to have an advantage in using XFFT for inter-station processing for the chosen parameters. If $N_\textrm{s}$ increases or if the cadence increases to $t_\textrm{acc}\lesssim 1$~ms, then EPIC becomes more efficient. 
\end{itemize}

@@

@@

% \subsection{LAMBDA-I}

>>>>

\author{S. Author}
\affiliation{Second Division, Organization, City, Pincode, State, Country}
% \alsoaffiliation{Joint first authors}

\author{T. Author}
\affiliation{Second Division, Organization, City, Pincode, State, Country}

\author{F.T. Author}
\affiliation{Fourth Division, Organization, City, Pincode, State, Country}

@@

\cite{TMS2017},
\cite{Otobe+1994,Tegmark+2009,Tegmark+2010,Zheng+2014},
\cite{Morales2011,Thyagarajan+2017},
\cite{Cornwell+2008},
\cite{Beardsley+2017},
\cite{Price2024},

@@

\section{Motivations}\label{sec:motivation}

This study is motivated by (a) the field of radio transient phenomena that span large ranges in time, and (b) large, modern radio interferometer arrays that access spatial scales spanning several orders of magnitude.

\subsection{Scientific: Time domain astronomy}\label{sec:time-domain-astronomy}

The observed phenomena of radio transients span orders of magnitude in timescales -- provide examples \cite{Pietka+2015}.

\subsection{Technical: Multi-scale, Hierarchical Aperture Arrays} \label{sec:multi-scale-arrays}

The specific aperture array examples we consider are the \texttt{SKA-low}, core of the SKA-low (\texttt{SKA-low-core}), \texttt{LAMBDA-I}, \texttt{CASPA} \citep{Luo+2024}, and \texttt{FarView}. The last three are hypothetical interferometers. 
% It is assumed that $N_{\textrm{a/s}_m}=N_\textrm{a/s}=256$ (same for each station), $D_\textrm{a}/\lambda=1$, and $D_\textrm{s}/\lambda=17.5$ for each of these arrays, where, the nominal wavelength chosen here is $\lambda=2$~m. The corresponding station filling factor, $f_\textrm{s}=N_\textrm{a/s}\,(D_\textrm{a}/D_\textrm{s})^2$, is 0.836 for all the arrays. ~\texttt{SKA-low-core}, ~\texttt{SKA-low}, ~\texttt{Aus-VLBA-low}, and ~\texttt{IC-VLBA-low} are taken to consist of $N_\textrm{s}=256, 512, 512,$ and 512 stations in an array diameter of $D_\textrm{A}/\lambda=525, 3.25\times 10^4, 1.95\times 10^6$, and $3.9\times 10^6$, respectively. The corresponding array filling factors, defined as $f_\textrm{A}=N_\textrm{s}\,(D_\textrm{s}/D_\textrm{A})^2$, are 0.284, $1.48\times 10^{-4}$, $4.12\times 10^{-8}$, and $1.03\times 10^{-8}$, respectively. 
These parameters for the chosen examples are summarised in Table~\ref{tab:array_params}.

@@

@@

Fast imaging involves processing data on multiple time scales. The channelisation of a voltage time stream of length $\delta t$ with a time resolution of $t_\textrm{s}$ will correspond to spectral channels of width $\delta\nu=1/\delta t$ and a bandwidth of $B=1/t_\textrm{s}$. The number of spectral channels is $N_\nu=B/\delta\nu=\delta t/t_\textrm{s}$. Images are expected to be produced at a cadence of $t_\textrm{acc}$, which is typically larger than $\delta t$ and smaller than the time scale on which calibration needs to be updated. In this paper, the nominal values chosen for $\delta t$ is 25~$\mu$s, and we explore the possibilities for $t_\textrm{acc}$ ranging from 100~$\mu$s to 10~s.

@@

Below, we provide a comparative view of the various architectures from the point of view of computational cost density for the range of array layouts and imaging cadence considered in this paper.

@@

Figure~\ref{fig:1D-coherent-compcost-a} shows a comparison of the compute cost densities in the discovery phase space of different multi-scale imaging architectures combining four options -- VBF, EPIC, XBF, and XFFT -- on the inter-station data with two options -- VBF and EPIC -- on the intra-station data for various array parameters and imaging cadences.

\begin{figure*}
\centering
\subfloat[][\texttt{LAMBDA-I} \label{fig:1D-coherent-compcost-LAMBDA-I}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_LAMBDA-I.pdf}
} \\
\subfloat[][\texttt{SKA1-Low-core} \label{fig:1D-coherent-compcost-LAMBDA-I}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_SKA1-low-core.pdf}
} \\
\subfloat[][\texttt{SKA1-Low} \label{fig:1D-coherent-compcost-LAMBDA-I}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_SKA1-low.pdf}
} 
\caption{One-dimensional slices of the computational cost density for (a) \texttt{LAMBDA-I}, (b) \texttt{SKA-low-core}, and (c) \texttt{SKA-low} using  coherent imaging at intra- and inter-station levels via VBF (gray) and EPIC (black) for intra-station processing, and VBF (dotted), EPIC (solid), XBF (dashed), and XFFT (dot-dashed) for inter-station processing of data. \label{fig:1D-coherent-compcost-a}}
\end{figure*}

\begin{figure*}\ContinuedFloat
\subfloat[][\texttt{CASPA} \label{fig:1D-coherent-compcost-CASPA}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_CASPA.pdf}
} \\
\subfloat[][\texttt{FarView} \label{fig:1D-coherent-compcost-FarView}]
{\includegraphics[width=\textwidth]
{figures/1D_coherent_per_pixel_compcost_analysis_FarView.pdf}
}
\caption{Contd. Same as above but for (d) \texttt{CASPA}, and (e) \texttt{FarView} \label{fig:1D-coherent-compcost-b}}
\end{figure*}

@@

The inter-station processing, however, varies significantly due to the diversity in the various inter-station array parameters and imaging cadence considered. For example, 
\begin{itemize}
    \item \texttt{LAMBDA-I} will benefit from deploying the correlator beamforming (XBF) architecture for inter-station processing. 
    \item \texttt{SKA1-low-core} will benefit from EPIC for the chosen parameters. However, if the number of stations, $N_\textrm{s}$, decreases,  the array diameter, $D_\textrm{A}$, increases or if the imaging cadence interval is slowed to $t_\textrm{acc} \gtrsim 5$~ms, the XFFT architecture starts to have an advantage over EPIC on the inter-station scale of data processing. 
    \item \texttt{SKA1-low} will benefit from the XFFT architecture for inter-station processing.
    \item \texttt{CASPA} will benefit from using the XBF architecture for inter-station processing due to significantly small number of stations.
    \item \texttt{FarView} appears to have an advantage in using XFFT for inter-station processing for the chosen parameters. If $N_\textrm{s}$ increases or if the cadence interval decreases to $t_\textrm{acc}\lesssim 1$~ms, then EPIC becomes more efficient. 
\end{itemize}

@@

The cadence of imaging is a key discriminator of the choice of imaging architectures. Here, we examine how the cadence interval determines the choice of imaging architecture for multi-scale arrays for intra- and inter-station processing. Table~\ref{tab:cadence} summarises the findings from Figures~\ref{fig:1D-incoherent-compcost} and \ref{fig:1D-coherent-compcost-b} based on the dependence of the computational cost densities of the different architectures and arrays on $t_\textrm{acc}$.

@@

\subsection{Other considerations}\label{sec:other-metrics}

Besides computational costs, there are other practical considerations in the deployment of an architecture such as the hardware's processing bandwidth, internal memory bandwidth management, input and output data rates, etc.\footnote{\url{https://github.com/epic-astronomy/Memos/blob/master/PDFs/009_EPIC_Code_Optimizations.md}}

\cite{Sokolowski+2024}

Post-processing steps like deconvolution and de-dispersion add to these costs.
